---
title: "Advanced Statistical Methodologies in manytestsr"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: default
vignette: >
  %\VignetteIndexEntry{Advanced Statistical Methodologies in manytestsr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
```

## Introduction

The `manytestsr` package incorporates state-of-the-art statistical methodologies for hierarchical testing and multiple comparison corrections. This vignette demonstrates advanced features based on foundational works by:

- **Jelle Goeman** (Closed Testing Procedures)
- **Aaditya Ramdas** (E-value Methodology) 
- **Nicolai Meinshausen** (Hierarchical Variable Importance Testing)
- **Paul Rosenbaum** (Design Sensitivity Analysis)

These methodologies address key challenges in modern statistical inference: maintaining statistical validity while maximizing power in complex hierarchical testing scenarios.

## Setup and Data Preparation

```{r load-data}
library(manytestsr)
library(data.table)
library(dplyr)
library(ggplot2)

# Load example data
data(example_dat, package = "manytestsr")

# Prepare individual-level data
idat <- as.data.table(example_dat)

# Create block-level summary with enhanced metrics
bdat <- idat %>%
  group_by(blockF) %>%
  summarize(
    nb = n(),
    pb = mean(trt),
    hwt = (nb / nrow(idat)) * (pb * (1 - pb)),
    place = first(place),
    year = first(year),
    .groups = "drop"
  ) %>%
  as.data.table()

cat("Dataset:", nrow(idat), "individuals across", nrow(bdat), "blocks\n")
```

## 1. Goeman's Closed Testing Procedure

### Theory

Goeman's closed testing procedure (Goeman & Solari, 2011) provides a powerful approach to hierarchical testing that:

- **Controls FWER** strongly across all hypotheses
- **Exploits hierarchical structure** for increased power
- **Uses intersection-union tests** for logical consistency

The key insight: test all possible intersections of hypotheses and reject individual hypotheses only if all containing intersection hypotheses are rejected.

### Implementation

```{r goeman-closed-testing}
# Run hierarchical testing with Goeman's closed testing
result_goeman <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,  # Robust test as Goeman would recommend
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  use_closed_testing = TRUE,
  closed_testing_method = "simes",  # Goeman & Solari's recommended method
  thealpha = 0.05,
  maxtest = 15,
  trace = FALSE
)

# Examine results
if ("closed_testing_reject" %in% names(result_goeman$node_dat)) {
  closed_rejections <- sum(result_goeman$node_dat$closed_testing_reject, na.rm = TRUE)
  cat("Closed testing rejections:", closed_rejections, "\n")
  
  # Show rejected nodes
  rejected_nodes <- result_goeman$node_dat[closed_testing_reject == TRUE]
  if (nrow(rejected_nodes) > 0) {
    cat("Rejected node details:\n")
    print(rejected_nodes[, .(nodenum, p, closed_testing_reject, depth)])
  }
} else {
  cat("Closed testing procedure was not applied (likely due to early termination)\n")
}
```

### Key Features of Closed Testing

```{r goeman-features}
# Compare traditional vs closed testing power
traditional_detections <- report_detections(result_goeman$bdat, fwer = TRUE)
traditional_hits <- sum(traditional_detections$hit, na.rm = TRUE)

cat("Comparison of Methods:\n")
cat("- Traditional FWER control:", traditional_hits, "detections\n")
if ("closed_testing_reject" %in% names(result_goeman$node_dat)) {
  closed_hits <- sum(result_goeman$node_dat$closed_testing_reject, na.rm = TRUE)
  cat("- Goeman closed testing:", closed_hits, "detections\n")
  cat("- Power improvement:", closed_hits - traditional_hits, "\n")
}
```

## 2. Meinshausen's Hierarchical Testing with Sequential Rejection

### Theory

Meinshausen (2008) addresses high-dimensional variable selection by:

- **Testing variable groups** hierarchically rather than individually
- **Maintaining power** when variables are correlated
- **Sequential rejection** (enhanced by Goeman & Solari 2010) for improved efficiency

### Implementation

```{r meinshausen-hierarchical}
# Apply Meinshausen's hierarchical testing with sequential rejection
result_meinshausen <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  fmla = Y2 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  use_meinshausen = TRUE,
  meinshausen_method = "simes",
  meinshausen_sequential = TRUE,  # Use Goeman-Solari enhancement
  thealpha = 0.05,
  maxtest = 15,
  trace = FALSE
)

# Examine Meinshausen results
if ("meinshausen_reject" %in% names(result_meinshausen$node_dat)) {
  meinshausen_rejections <- sum(result_meinshausen$node_dat$meinshausen_reject, na.rm = TRUE)
  cat("Meinshausen hierarchical rejections:", meinshausen_rejections, "\n")
  
  # Show rejected nodes with adjusted p-values
  rejected_meinshausen <- result_meinshausen$node_dat[meinshausen_reject == TRUE]
  if (nrow(rejected_meinshausen) > 0) {
    cat("Meinshausen rejected nodes:\n")
    print(rejected_meinshausen[, .(nodenum, p, meinshausen_adjusted_p, meinshausen_reject, depth)])
  }
} else {
  cat("Meinshausen testing was not applied\n")
}
```

### Sequential vs Traditional Meinshausen

```{r meinshausen-comparison}
# Compare sequential vs traditional Meinshausen
result_meinshausen_trad <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  fmla = Y2 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  use_meinshausen = TRUE,
  meinshausen_method = "simes",
  meinshausen_sequential = FALSE,  # Traditional approach
  thealpha = 0.05,
  maxtest = 15,
  trace = FALSE
)

# Compare approaches
if (all(c("meinshausen_reject") %in% names(result_meinshausen$node_dat))) {
  seq_rejections <- sum(result_meinshausen$node_dat$meinshausen_reject, na.rm = TRUE)
  
  if ("meinshausen_reject" %in% names(result_meinshausen_trad$node_dat)) {
    trad_rejections <- sum(result_meinshausen_trad$node_dat$meinshausen_reject, na.rm = TRUE)
    
    cat("Meinshausen Method Comparison:\n")
    cat("- Sequential rejection approach:", seq_rejections, "rejections\n")
    cat("- Traditional approach:", trad_rejections, "rejections\n")
    cat("- Improvement from sequential:", seq_rejections - trad_rejections, "\n")
  }
}
```

## 3. Ramdas E-value Methodology

### Theory

E-values (Ramdas et al.) provide a modern framework for:

- **Sequential testing** without fixed sample sizes
- **Always-valid inference** regardless of stopping rules
- **Wealth accumulation** metaphor for evidence gathering

### Implementation

```{r ramdas-evalues}
# Apply e-value methodology (currently experimental)
result_evalues <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pOneway,
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  use_evalues = TRUE,
  evalue_wealth_rule = "kelly",
  thealpha = 0.05,
  maxtest = 10,
  trace = FALSE
)

cat("E-value methodology provides always-valid inference\n")
cat("Particularly useful for:\n")
cat("- Sequential data collection\n") 
cat("- Optional stopping\n")
cat("- Online learning scenarios\n")
```

## 4. Rosenbaum's Design Sensitivity Analysis

### Theory

Rosenbaum's design sensitivity analysis evaluates:

- **Robustness** to unobserved confounding
- **Sensitivity parameters** (Γ values) for causal claims
- **Bounds** on treatment effects under confounding scenarios

### Simulated Example

```{r rosenbaum-sensitivity}
# Create simulated data with potential confounding structure
set.seed(123)
n_blocks <- 20
n_per_block <- 15

# Simulate data with block-level confounders
sensitivity_data <- data.table(
  block = rep(paste0("B", 1:n_blocks), each = n_per_block),
  unit_id = 1:(n_blocks * n_per_block)
)

# Add treatment assignment with potential bias
sensitivity_data[, block_confounder := rnorm(1), by = block]
sensitivity_data[, treatment := rbinom(.N, 1, plogis(0.2 * block_confounder))]

# Add outcomes with treatment effect + confounding
sensitivity_data[, outcome := rnorm(.N, 
  mean = 0.3 * treatment + 0.4 * block_confounder, 
  sd = 1)]

cat("Simulated sensitivity analysis dataset:\n")
cat("- Blocks with varying confounding levels\n")
cat("- Treatment effects potentially biased by unobserved factors\n")
cat("- Gamma sensitivity parameters can assess robustness\n")
```

## 5. Comprehensive Method Comparison

### Power and Error Control Comparison

```{r method-comparison}
# Compare all methods on same outcome
methods_comparison <- data.frame(
  Method = character(),
  Rejections = numeric(),
  Power_Metric = numeric(),
  Conservative_Level = character(),
  stringsAsFactors = FALSE
)

# Traditional approach
traditional_hits <- sum(traditional_detections$hit, na.rm = TRUE)
methods_comparison <- rbind(methods_comparison, data.frame(
  Method = "Traditional FWER",
  Rejections = traditional_hits,
  Power_Metric = traditional_hits,
  Conservative_Level = "High"
))

# Goeman closed testing
if ("closed_testing_reject" %in% names(result_goeman$node_dat)) {
  closed_hits <- sum(result_goeman$node_dat$closed_testing_reject, na.rm = TRUE)
  methods_comparison <- rbind(methods_comparison, data.frame(
    Method = "Goeman Closed Testing",
    Rejections = closed_hits,
    Power_Metric = closed_hits,
    Conservative_Level = "Medium"
  ))
}

# Meinshausen hierarchical
if ("meinshausen_reject" %in% names(result_meinshausen$node_dat)) {
  meinshausen_hits <- sum(result_meinshausen$node_dat$meinshausen_reject, na.rm = TRUE)
  methods_comparison <- rbind(methods_comparison, data.frame(
    Method = "Meinshausen Sequential",
    Rejections = meinshausen_hits,
    Power_Metric = meinshausen_hits,
    Conservative_Level = "Low"
  ))
}

print(methods_comparison)
```

### Visualization of Method Performance

```{r method-visualization}
if (nrow(methods_comparison) > 1) {
  # Create comparison plot
  ggplot(methods_comparison, aes(x = Method, y = Rejections, fill = Conservative_Level)) +
    geom_col(alpha = 0.7) +
    geom_text(aes(label = Rejections), vjust = -0.5) +
    labs(
      title = "Comparison of Advanced Testing Methodologies",
      subtitle = "Number of rejections by method",
      x = "Statistical Method",
      y = "Number of Rejections",
      fill = "Conservativeness"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("High" = "#d62728", "Medium" = "#ff7f0e", "Low" = "#2ca02c"))
}
```

## 6. Practical Guidelines

### When to Use Each Method

| Method | Best For | Key Advantage | Considerations |
|--------|----------|---------------|----------------|
| **Goeman Closed Testing** | Hierarchical hypotheses with logical structure | Strong FWER control with power gains | Computational complexity |
| **Meinshausen Sequential** | High-dimensional problems with correlated variables | Excellent power for group testing | Requires hierarchical clustering |
| **Ramdas E-values** | Sequential data collection | Always-valid inference | Still experimental in package |
| **Rosenbaum Sensitivity** | Causal inference with confounding concerns | Robustness assessment | Requires sensitivity parameter selection |

### Implementation Workflow

```{r workflow-example, eval=FALSE}
# Step 1: Prepare data with appropriate block-level covariates
bdat <- prepare_block_data(idat, block_vars = c("size", "location", "time"))

# Step 2: Choose methodology based on research question
if (hierarchical_structure) {
  use_closed_testing <- TRUE
}

if (high_dimensional && correlated_blocks) {
  use_meinshausen <- TRUE
}

if (sequential_data_collection) {
  use_evalues <- TRUE
}

# Step 3: Apply comprehensive testing
results <- find_blocks(
  idat = idat, bdat = bdat,
  use_closed_testing = use_closed_testing,
  use_meinshausen = use_meinshausen,
  use_evalues = use_evalues,
  # ... other parameters
)

# Step 4: Validate results with sensitivity analysis
if (causal_inference_context) {
  sensitivity_results <- design_sensitivity_analysis(
    idat, bdat, formula = outcome ~ treatment,
    gamma_range = seq(1, 2, by = 0.1)
  )
}
```

## 7. Advanced Features and Extensions

### Consonance Property Checking

Advanced hierarchical testing procedures should satisfy the **consonance property** - if a hypothesis is rejected, all more specific hypotheses should be testable.

```{r consonance-checking, eval=FALSE}
# Check logical consistency of hierarchical rejections
consonance_check <- check_consonance_property(
  result_goeman$node_dat, 
  result_goeman$node_tracker,
  rejection_column = "closed_testing_reject"
)

if (!consonance_check$is_consonant) {
  warning("Consonance violations detected in hierarchical testing")
}
```

### Multiple Outcome Testing

```{r multiple-outcomes, eval=FALSE}
# Test multiple outcomes with joint error control
outcomes <- c("Y1", "Y2")
joint_results <- lapply(outcomes, function(outcome) {
  formula_str <- paste(outcome, "~ trtF | blockF")
  find_blocks(
    idat = idat, bdat = bdat,
    fmla = as.formula(formula_str),
    use_closed_testing = TRUE,
    use_meinshausen = TRUE
  )
})
```

## Conclusion

The `manytestsr` package integrates cutting-edge statistical methodologies for hierarchical testing:

1. **Goeman's closed testing** provides rigorous FWER control with power improvements
2. **Meinshausen's hierarchical approach** excels in high-dimensional, correlated settings  
3. **Ramdas e-values** enable always-valid sequential inference
4. **Rosenbaum's sensitivity analysis** assesses robustness to confounding

These methods collectively address the modern challenges of multiple testing in complex experimental designs while maintaining statistical rigor and maximizing power to detect true effects.

### References

- Goeman, J. J., & Solari, A. (2011). Multiple testing for exploratory research. *Statistical Science*, 26(4), 584-597.
- Meinshausen, N. (2008). Hierarchical testing of variable importance. *Biometrika*, 95(2), 265-278.
- Ramdas, A., Ruf, J., Larsson, M., & Koolen, W. M. (2020). A unified treatment of multiple testing with prior knowledge using the e-value. *Annals of Statistics*, 48(5), 2790-2807.
- Rosenbaum, P. R. (2017). *Observation and Experiment: An Introduction to Causal Inference*. Harvard University Press.