---
title: "Getting Started with manytestsr"
format:
  html:
    toc: true
    toc-depth: 2
    code-fold: false
vignette: >
  %\VignetteIndexEntry{Getting Started with manytestsr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
# Some tests below simulate permutation distributions and one splitter
# falls back to random assignment on ties; fix the seed so the rendered
# numbers are stable across site rebuilds.
set.seed(20260812)
```

## What is Hierarchical Testing?

The `manytestsr` package implements hierarchical testing procedures for detecting treatment effects across multiple experimental blocks. Instead of testing each block independently (which inflates Type I error) or testing everything together (which reduces power), hierarchical testing:

1. **Splits blocks** into groups based on similarity or pre-specified structure
2. **Tests groups** at each level of the hierarchy  
3. **Controls error rates** while maintaining power to detect heterogeneous effects

This is especially useful for:
- Multi-site experiments with varying treatment effects
- A/B tests across different user segments  
- Clinical trials with multiple centers
- Any setting with clustered/blocked experimental units

## Quick Start

### Load Package and Data

```{r quickstart}
library(manytestsr)
library(data.table)
library(dplyr)

# Load built-in example data
data(example_dat, package = "manytestsr")

# Prepare individual-level data
idat <- as.data.table(example_dat)

# Create block-level summary
bdat <- idat %>%
  group_by(blockF) %>%
  summarize(
    nb = n(),                    # Block size
    pb = mean(trt),             # Proportion treated
    hwt = (nb / nrow(idat)) * (pb * (1 - pb)),  # Harmonic mean weight
    .groups = "drop"
  ) %>%
  as.data.table()

print(paste("Data:", nrow(idat), "individuals in", nrow(bdat), "blocks"))
```

### Run Hierarchical Testing

```{r basic-example}
# Basic hierarchical testing
results <- find_blocks(
  idat = idat,                    # Individual-level data
  bdat = bdat,                    # Block-level data
  blockid = "blockF",             # Block identifier column
  splitfn = splitCluster,         # How to split blocks (k-means clustering)
  pfn = pOneway,                  # Statistical test to use (t-test)
  fmla = Y1 ~ trtF | blockF,     # Formula: outcome ~ treatment | block
  splitby = "hwt",               # Variable to guide splitting
  parallel = "no",               # Disable parallel processing for demo
  thealpha = 0.05                # Overall error rate
)

print(paste("Testing created", nrow(results$node_dat), "nodes in the tree"))
```

### Find Significant Effects

```{r detect-effects}
# Identify blocks with detected treatment effects
detections <- report_detections(results$bdat, fwer = TRUE, alpha = 0.05)

# Summary (hit is never NA, so plain sums work)
cat("Results Summary:\n")
cat("- Total blocks tested:", nrow(detections), "\n")
cat("- Blocks with detected effects:", sum(detections$hit), "\n")
cat("- Detection rate:", round(mean(detections$hit) * 100, 1), "%\n")

# Detections come in two kinds (see prose below)
cat("\nDetections by type:\n")
print(table(detections$hit_type))

# Show detected blocks if any found
if (any(detections$hit)) {
  sig_blocks <- detections[hit == TRUE, .(blockF, hit_type, pfinalb, group_p)]
  cat("\nDetected blocks:\n")
  print(sig_blocks)
}
```

The `hit_type` column separates two kinds of findings. A `"single"` hit
means the block's own test rejected: the procedure localized an effect to
that specific block. A `"group"` hit means the procedure rejected the
block's parent group but no test within that group rejected: we know the
group contains an effect somewhere, yet cannot say which block carries it.
For group hits, `group_p` reports the rejecting parent's p-value while
`pfinalb` shows the (non-significant) result of the block's own test, so
the two columns together record both what was found and what was not.
Blocks with `hit_type == "none"` include those under a rejected parent
whose rejection a sibling's own detection already explains.

## Key Components

### Splitting Functions

Choose how to divide blocks at each step:

```{r splitting-functions, eval=FALSE}
# Cluster-based splitting (most common)
splitCluster      # Groups similar blocks using k-means

# Pre-specified hierarchical splitting  
splitSpecifiedFactor  # Follows predefined hierarchy (e.g., state > district > school)

# Leave-one-out splitting
splitLOO          # Focuses on largest/most powerful blocks first

# Equal-sum splitting
splitEqualApprox  # Balances total size/weight across groups
```

### Test Functions

Choose the statistical test:

```{r test-functions, eval=FALSE}
pOneway       # T-tests (assumes normality)
pIndepDist    # Distance-based tests (robust, recommended)
pWilcox       # Wilcoxon rank-sum tests (ordinal outcomes)
```

### Error Control

The recommended setting is a fixed alpha. The gated top-down testing
limits which hypotheses are ever tested: a group is split and its
children tested only after the group's own test rejects. Whether that
gating alone controls the familywise error rate (FWER) depends on the
design; `compute_error_load()` (fed a headcount column such as `nb`, not
weights) diagnoses whether your design needs depth-adjusted alpha levels
beyond the gating.

```{r error-control, eval=FALSE}
# Fixed alpha -- the default and recommended setting
alphafn = NULL, thealpha = 0.05
```

The package also wraps sequential alpha procedures from the `onlineFDR`
package (`alpha_investing`, `alpha_saffron`, `alpha_addis`). These treat
the tree's p-values as a flat stream; their guarantees are proven for
that stream setting, not for gated tree-structured testing, so we do not
recommend them and plan to deprecate them (see the package README).

## Example: Robust Distance-Based Testing

```{r robust-example}
results_robust <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,           # Distance-based test (robust)
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no"
)

robust_detections <- report_detections(results_robust$bdat)
cat("Robust approach detections:", sum(robust_detections$hit), "\n")
```

The experimental sequential procedures are invoked as follows. We show
the call without running it: as noted above, their error control is not
established for tree-structured testing.

```{r fdr-example, eval=FALSE}
results_seq <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  alphafn = alpha_investing,  # Experimental: no proven guarantee here
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  thealpha = 0.05,
  thew0 = 0.049               # Starting "wealth"
)
```

## Visualizing Results

### Tree Structure

```{r visualization, fig.width=8, fig.height=6}
library(ggraph)
library(ggplot2)

# Create tree visualization
tree_data <- make_results_tree(results, block_id = "blockF")
tree_plot <- make_results_ggraph(tree_data$graph)

# Display the tree
tree_plot + 
  labs(title = "Hierarchical Testing Tree") +
  theme_void()
```

### Results Summary

```{r results-table}
# Create summary table
summary_table <- tree_data$test_summary
if(!is.null(summary_table) && is.data.frame(summary_table) && nrow(summary_table) > 0) {
  cat("Test Summary:\n")
  print(summary_table)
} else {
  cat("Tree structure (nodes by depth):\n")
  if(!is.null(tree_data$nodes) && is.data.frame(tree_data$nodes)) {
    print(tree_data$nodes[, .N, by = depth])
  } else {
    cat("Tree data structure available but not displayed in simple format.\n")
  }
}
```

## Best Practices

### 1. Data Preparation
- Ensure treatment assignment is balanced within blocks
- Include relevant block-level covariates for splitting
- Calculate appropriate power weights (harmonic mean weights work well)

### 2. Method Selection
- **Start with**: `splitCluster` + `pIndepDist` with a fixed alpha (the default)
- **For pre-specified hierarchies**: Use `splitSpecifiedFactor`
- **For robustness**: Always consider `pIndepDist`
- **For multiple outcomes**: Add local p-value adjustment (e.g. `local_simes`)

### 3. Interpretation
- Focus on blocks identified as significant
- Consider effect sizes, not just p-values
- Validate findings with additional data if possible

## Next Steps

The **Hierarchical Testing with manytestsr** vignette walks through a
complete analysis: alternative splitting strategies, detection reporting
under FWER control, and visualization of the testing tree.

By testing groups of blocks before individual blocks, `find_blocks()`
spends few tests at the top of the tree and descends only into groups
whose tests rejected, retaining power to find the blocks where treatment
effects concentrate. Whether that gating alone controls the familywise
error rate depends on the design; `compute_error_load()` diagnoses when
depth-adjusted alpha levels are needed.
