---
title: "Hierarchical Testing with manytestsr"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    theme: default
vignette: >
  %\VignetteIndexEntry{Hierarchical Testing with manytestsr}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 6
)
# Some tests below simulate permutation distributions and one splitter
# falls back to random assignment on ties; fix the seed so the rendered
# numbers are stable across site rebuilds.
set.seed(20260812)
```

## Introduction

The `manytestsr` package implements hierarchical testing procedures for detecting treatment effects across multiple experimental blocks while controlling error rates. This approach is particularly useful when you have:

- Multiple experimental units organized in blocks
- Heterogeneous treatment effects across blocks
- Need to identify which specific blocks show effects
- Want to control family-wise error rate (FWER) or false discovery rate (FDR)

This vignette walks through the complete workflow from data preparation to results interpretation.

## Loading the Package and Data

```{r load-packages}
library(manytestsr)
library(data.table)
library(dplyr)
library(ggplot2)
library(ggraph)

# Load example data
data(example_dat, package = "manytestsr")
head(example_dat)
```

The example dataset contains:

- `id`: Individual unit identifier
- `blockF`: Block (cluster) factor
- `trtF`: Treatment assignment factor (0 = control, 1 = treatment)
- `Y1`, `Y2`: Outcome variables
- `place`: Location identifier
- `year`: Time identifier
- `place_year_block`: Hierarchical grouping variable

## Data Preparation

The hierarchical testing approach requires both individual-level data (`idat`) and block-level summaries (`bdat`):

```{r data-prep}
# Individual-level data is already in the right format
idat <- as.data.table(example_dat)
print(paste("Number of individuals:", nrow(idat)))
print(paste("Number of blocks:", length(unique(idat$blockF))))

# Create block-level dataset with key variables
bdat <- idat %>%
  group_by(blockF) %>%
  summarize(
    # Sample size
    nb = n(),
    # Proportion treated
    pb = mean(trt),
    # Harmonic mean weight (for testing power)
    hwt = (nb / nrow(idat)) * (pb * (1 - pb)),
    # Block characteristics
    place = unique(place),
    year = unique(year),
    place_year_block = factor(unique(place_year_block)),
    .groups = "drop"
  ) %>%
  as.data.table()

head(bdat)
```

Key block-level variables:

- **`hwt`** (harmonic mean weight): Measures testing power for each block
- **`nb`** (block size): Number of units in each block
- **`pb`** (proportion treated): Treatment assignment rate
- **`place_year_block`**: Hierarchical factor for pre-specified splits

## Basic Hierarchical Testing

### Using Cluster-Based Splitting

The most common approach uses clustering to split blocks based on a continuous variable:

```{r basic-clustering}
# Run hierarchical testing with cluster-based splitting
results_cluster <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster, # Split using k-means clustering
  pfn = pOneway, # Use t-tests
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt", # Split based on harmonic mean weights
  parallel = "no", # Disable parallel processing for demo
  trace = TRUE, # Show split progression
  thealpha = 0.05
)

# Examine the structure
str(results_cluster, max.level = 1)
```

### Results Overview

```{r results-summary}
# Block-level results
cat("Block-level results structure:\n")
cat("Number of blocks:", nrow(results_cluster$bdat), "\n")
cat("Variables:", names(results_cluster$bdat), "\n\n")

# Node-level results
cat("Node-level results structure:\n")
cat("Number of nodes:", nrow(results_cluster$node_dat), "\n")
cat("Depth levels:", sort(unique(results_cluster$node_dat$depth)), "\n")
```

## Alternative Splitting Strategies

### Pre-specified Hierarchical Splitting

When you have natural hierarchical structure, use factor-based splitting:

```{r hierarchical-splitting}
# Use pre-specified hierarchical splits
results_hierarchical <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitSpecifiedFactor,
  pfn = pIndepDist, # Use distance-based test
  fmla = Y2 ~ trtF | blockF,
  splitby = "place_year_block",
  parallel = "no",
  trace = TRUE,
  thealpha = 0.05
)

print(paste(
  "Hierarchical approach found",
  nrow(results_hierarchical$node_dat), "nodes"
))
```

### Leave-One-Out Splitting

Focus testing on largest blocks first:

```{r loo-splitting}
# Leave-one-out approach
results_loo <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitLOO,
  pfn = pOneway,
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt", # Focus on blocks with highest power
  parallel = "no",
  thealpha = 0.05
)

print(paste(
  "LOO approach found",
  nrow(results_loo$node_dat), "nodes"
))
```

## Sequential Alpha Procedures (Experimental)

The package wraps sequential alpha procedures from the `onlineFDR`
package (`alpha_investing`, `alpha_saffron`, `alpha_addis`). These
procedures treat the tree's p-values as a flat stream, and their
guarantees are proven for that stream setting, not for gated
tree-structured testing. We do not recommend them for applied work and
plan to deprecate them (see the package README). We run one here only to
show how the per-node alpha levels adapt as testing proceeds.

```{r alpha-investing}
# Alpha investing: experimental in the tree setting (see caution above)
results_fdr <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  alphafn = alpha_investing, # Sequential alpha adjustment
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  thealpha = 0.05,
  thew0 = 0.049 # Starting wealth
)

# Compare alpha levels across approaches
alpha_comparison <- data.frame(
  Node = 1:min(nrow(results_cluster$node_dat), nrow(results_fdr$node_dat)),
  Fixed_Alpha = results_cluster$node_dat$a[1:min(
    nrow(results_cluster$node_dat),
    nrow(results_fdr$node_dat)
  )],
  Adaptive_Alpha = results_fdr$node_dat$a[1:min(
    nrow(results_cluster$node_dat),
    nrow(results_fdr$node_dat)
  )]
)

print("Alpha level comparison:")
head(alpha_comparison)
```

## Detecting Significant Effects

### Using FWER Control

```{r detect-effects-fwer}
# Detect significant blocks using FWER control
detections_fwer <- report_detections(
  results_cluster$bdat,
  fwer = TRUE,
  alpha = 0.05,
  blockid = "blockF"
)

# Summary of detections (hit is never NA, so plain sums work)
cat("FWER Results:\n")
cat("Total blocks:", nrow(detections_fwer), "\n")
cat("Detected blocks:", sum(detections_fwer$hit), "\n")
cat(
  "Detection rate:",
  round(mean(detections_fwer$hit) * 100, 1), "%\n\n"
)

# Detections by type: "single" = the block's own test rejected;
# "group" = the block's parent group was rejected but the effect could
# not be attributed to specific blocks within it
cat("Detections by type:\n")
print(table(detections_fwer$hit_type))

# Show detected blocks
if (any(detections_fwer$hit)) {
  significant_blocks <- detections_fwer[
    hit == TRUE,
    .(blockF, hit_type, pfinalb, group_p, fin_nodenum)
  ]
  print("Detected blocks:")
  print(significant_blocks)
}
```

The `hit_type` column keeps the two kinds of findings separate. A
`"single"` hit localizes an effect to that block: its own test rejected.
A `"group"` hit records a coarser finding: the block's parent group was
rejected while no test inside the group rejected, so the effect sits
somewhere in the group without being attributed to particular blocks.
For group hits, `group_p` carries the rejecting parent's p-value; the
block's own `pfinalb` stays above alpha, which is exactly why the finding
remains at the group level. Blocks under a rejected parent do not inherit
coverage when a sibling's own detection already explains the parent's
rejection; those blocks show `hit_type == "none"`.

### Detections from the Experimental Sequential Run

The alpha-investing run from the previous section can be summarized the
same way. The caution given there applies to these counts too: no proven
error-rate guarantee attaches to them in the tree setting.

```{r detect-effects-fdr}
detections_fdr <- report_detections(
  results_fdr$bdat,
  fwer = FALSE, # Use each block's final-depth p rather than the path maximum
  alpha = 0.05
)

cat("Alpha-investing (experimental) results:\n")
cat("Detected blocks:", sum(detections_fdr$hit), "\n")
cat(
  "Detection rate:",
  round(mean(detections_fdr$hit) * 100, 1), "%\n"
)
```

## Visualizing Results

### Tree Structure Visualization

```{r tree-visualization, fig.width=10, fig.height=8}
# Create tree structure for visualization
tree_results <- make_results_tree(
  results_cluster,
  block_id = "blockF",
  node_label = "hwt"
)

# Create the graph visualization
tree_plot <- make_results_ggraph(tree_results$graph, remove_na_p = TRUE)

# Customize the plot
tree_plot_styled <- tree_plot +
  labs(
    title = "Hierarchical Testing Results Tree",
    subtitle = "Nodes colored by p-values, size by block weight",
    caption = "Red nodes indicate significant results"
  ) +
  theme_void() +
  theme(
    plot.title = element_text(size = 14, hjust = 0.5),
    plot.subtitle = element_text(size = 12, hjust = 0.5),
    plot.caption = element_text(size = 10, hjust = 0.5)
  )

print(tree_plot_styled)
```

### Detection Summary Plot

```{r detection-plot, fig.width=10, fig.height=6}
# Compare detection rates across methods
detection_summary <- data.frame(
  Method = c(
    "FWER (Cluster)", "Alpha investing (experimental)",
    "FWER (Hierarchical)", "FWER (LOO)"
  ),
  Detections = c(
    sum(detections_fwer$hit),
    sum(detections_fdr$hit),
    sum(report_detections(results_hierarchical$bdat, fwer = TRUE)$hit),
    sum(report_detections(results_loo$bdat, fwer = TRUE)$hit)
  ),
  Total_Blocks = c(
    nrow(detections_fwer),
    nrow(detections_fdr),
    nrow(results_hierarchical$bdat),
    nrow(results_loo$bdat)
  )
)

detection_summary$Detection_Rate <- detection_summary$Detections / detection_summary$Total_Blocks

# Create comparison plot
ggplot(detection_summary, aes(x = Method, y = Detection_Rate, fill = Method)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = paste0(Detections, "/", Total_Blocks)),
    vjust = -0.5, size = 3
  ) +
  labs(
    title = "Detection Rates Across Testing Methods",
    y = "Proportion of Blocks with Detected Effects",
    x = "Testing Method"
  ) +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )
```

## Interpreting P-values and Alpha Levels

### P-value Progression Through Tree

```{r pvalue-analysis}
# Examine p-value patterns
pvalue_data <- results_cluster$node_dat[, .(
  nodenum,
  depth,
  p,
  a,
  testable,
  nodesize
)]

# Show how p-values change with depth
pvalue_summary <- pvalue_data[!is.na(p), .(
  Mean_P = mean(p),
  Median_P = median(p),
  Mean_Alpha = mean(a),
  N_Nodes = .N
), by = depth]

print("P-value progression by tree depth:")
print(pvalue_summary)
```

### Statistical Power Analysis

```{r power-analysis, fig.width=10, fig.height=6}
# Analyze relationship between block characteristics and detection
power_data <- merge(
  detections_fwer[, .(blockF, hit, pfinalb)],
  bdat[, .(blockF, hwt, nb, pb)],
  by = "blockF"
)

# Power vs. block size
p1 <- ggplot(power_data, aes(x = nb, y = -log10(pfinalb), color = hit)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red")) +
  labs(
    title = "Detection by Block Size",
    x = "Block Size (number of units)",
    y = "-log10(p-value)",
    color = "Detected"
  ) +
  theme_minimal()

# Power vs. harmonic mean weight
p2 <- ggplot(power_data, aes(x = hwt, y = -log10(pfinalb), color = hit)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_manual(values = c("FALSE" = "blue", "TRUE" = "red")) +
  labs(
    title = "Detection by Harmonic Mean Weight",
    x = "Harmonic Mean Weight",
    y = "-log10(p-value)",
    color = "Detected"
  ) +
  theme_minimal()

# Combine plots
library(patchwork)
p1 + p2 + plot_layout(ncol = 2)
```

## Advanced Usage: Multiple Outcomes

### Testing Multiple Outcomes

```{r multiple-outcomes}
# Test both outcomes with local p-value adjustment
results_multi_Y1 <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  local_adj_p_fn = local_simes, # Simes adjustment within nodes
  fmla = Y1 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  thealpha = 0.05
)

results_multi_Y2 <- find_blocks(
  idat = idat,
  bdat = bdat,
  blockid = "blockF",
  splitfn = splitCluster,
  pfn = pIndepDist,
  local_adj_p_fn = local_simes,
  fmla = Y2 ~ trtF | blockF,
  splitby = "hwt",
  parallel = "no",
  thealpha = 0.05
)

# Compare results
multi_comparison <- data.frame(
  Outcome = c("Y1", "Y2"),
  Nodes = c(nrow(results_multi_Y1$node_dat), nrow(results_multi_Y2$node_dat)),
  Detections = c(
    sum(report_detections(results_multi_Y1$bdat)$hit),
    sum(report_detections(results_multi_Y2$bdat)$hit)
  )
)

print("Multiple outcome comparison:")
print(multi_comparison)
```

## Summary and Best Practices

### Key Takeaways

1. **Splitting Strategy Choice**:
   - **Cluster-based (`splitCluster`)**: Good for continuous block characteristics
   - **Hierarchical (`splitSpecifiedFactor`)**: Use when you have natural hierarchical structure
   - **Leave-one-out (`splitLOO`)**: Focus on highest-power blocks first

2. **Error Rate Control**:
   - **Fixed alpha**: the recommended default. Use `compute_error_load()`
     (with a headcount `blocksize` such as `nb`, not weights) to check
     whether the gated procedure needs depth-adjusted alpha levels for
     your design
   - **Sequential procedures** (`alpha_investing`, `alpha_saffron`,
     `alpha_addis`): experimental in the tree setting -- their guarantees
     are proven for flat streams of p-values, not for gated
     tree-structured testing, and they are slated for deprecation

3. **Test Function Selection**:
   - **`pOneway`**: Standard t-tests, good for normal outcomes
   - **`pIndepDist`**: Distance-based tests, robust to distributions
   - **`pWilcox`**: Rank-based tests, good for ordinal outcomes

### Recommended Workflow

```{r workflow-demo, eval=FALSE}
# 1. Prepare data: build the block-level summary (nb, pb, hwt) from the
#    individual-level data, as in the Data Preparation section above
idat <- your_individual_data
bdat <- your_block_level_summary

# 2. Choose approach based on data structure
if (have_hierarchical_structure) {
  splitfn <- splitSpecifiedFactor
  splitby <- "hierarchical_variable"
} else {
  splitfn <- splitCluster
  splitby <- "power_variable" # e.g., block size or weights
}

# 3. Run hierarchical testing with fixed alpha (the recommended default)
results <- find_blocks(
  idat = idat,
  bdat = bdat,
  splitfn = splitfn,
  pfn = pIndepDist, # Robust choice
  splitby = splitby,
  thealpha = 0.05
)

# 4. Detect effects; hit_type separates single-block from group findings
detections <- report_detections(results$bdat, fwer = TRUE)

# 5. Visualize results
tree <- make_results_tree(results, block_id = "block_variable")
plot <- make_results_ggraph(tree$graph)
```

### Performance Considerations

- Use `parallel = "multicore"` for faster computation on multi-core systems
- Set appropriate `simthresh` to balance accuracy vs. speed
- Consider `maxtest` to limit tree depth in very large datasets
- Use `trace = TRUE` during development to monitor progress

By testing groups of blocks before individual blocks, `find_blocks()` spends few tests at the top of the tree and descends only into groups whose tests rejected, retaining power to find the blocks where treatment effects concentrate. Whether that gating alone controls the familywise error rate depends on the design; `compute_error_load()` diagnoses when depth-adjusted alpha levels are needed. The `hit_type` column in `report_detections()` then records how far each finding was localized: to a single block, or only to a group.
